---
title: "The useage of GeoTcgaData"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GeoTcgaData}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
--------

## Authors
Erqiang Hu

College of Bioinformatics Science and Technology, Harbin Medical University

## Installation

Get the released version from CRAN:

``` {r, eval=FALSE, message=FALSE, warning=FALSE}
install.packages("GeoTcgaData")
```

Or the development version from github:

```{r, eval=FALSE, message=FALSE, warning=FALSE}
if(!requireNamespace("devtools", quietly = TRUE))
    install.packages("devtools")
devtools::install_github("huerqiang/GeoTcgaData")
```


## Introduction
GEO and TCGA provide us with a wealth of data, such as RNA-seq, DNA Methylation, single nucleotide Variation and Copy number variation data. It's easy to download data from TCGA using the  gdc tool or `TCGAbiolinks`, and some software provides organized TCGA data, such as [UCSC Xena](http://xena.ucsc.edu/)，but processing these data into a format suitable for bioinformatics  analysis requires more work. This R package was developed to handle these data.

```{r}
library(GeoTcgaData)
```

## Common operations on GeoTcgaData


This is a basic example which shows you how to solve a common problem:

## RNA-seq data integration and differential gene extraction

It is convenient to use [`TCGAbiolinks`](http://www.bioconductor.org/packages/release/bioc/vignettes/TCGAbiolinks/inst/doc/analysis.html)  or [`GDCRNATools`](https://bioconductor.org/packages/GDCRNATools/) to download and analysis Gene expression data.  `TCGAbiolinks` use `edgeR` package to do differential expression analysis, while `GDCRNATools` can implement three most commonly used methods: limma, edgeR , and DESeq2 to identify differentially expressed  genes (DEGs).

However, unlike the chip data, the RNA-seq data had one [bias](https://pubmed.ncbi.nlm.nih.gov/20132535/): the longer the gene, the more likely it was to be  identified as a differential gene, [while there was no such trend in the chip data](https://pubmed.ncbi.nlm.nih.gov/19371405/).  This is because in RNA-seq data a long gene has more reads mapping to it compared to a short gene of similar expression,  and most of the statistical methods used to detect differential expression  have stronger detection ability for genes with more reads. Therefore, we need to correct the gene length bias in downstream analysis  such as enrichment analysis.

[GOseq](http://bioconductor.org/packages/goseq/) based on [Wallenius' noncentral hypergeometric distribution](https://en.wikipedia.org/wiki/Wallenius%27_noncentral_hypergeometric_distribution) can effectively correct the gene length deviation in enrichment analysis. However, its algorithm can not directly correct the deviation of the expression profile, and its results can not be used for GSEA enrichment analysis. [CQN](http://www.bioconductor.org/packages/cqn/) present a [normalization algorithm](https://pubmed.ncbi.nlm.nih.gov/22285995/) to correct systematic biases(gene length bias and [GC-content bias](https://pubmed.ncbi.nlm.nih.gov/22177264/)), whose result can be seamlessly docked with downstream difference analysis  software such as  DESeq2 and edgeR. 

Here we use `TCGAbiolinks` to download RNA-seq data, use `CQN` to correct gene  length bias and GC content bias, and then use `DESeq2` for difference  analysis.

use `TCGAbiolinks` to download TCGA data

```{r, eval=FALSE, message=FALSE, warning=FALSE}
# download RNA-seq data
library(TCGAbiolinks)
                
query <- GDCquery(project = "TCGA-ACC",
                  data.category = "Transcriptome Profiling",
                  data.type = "Gene Expression Quantification", 
                  workflow.type = "HTSeq - Counts")
                  
GDCdownload(query, method = "api", files.per.chunk = 3, 
    directory = Your_Path)

dataRNA <- GDCprepare(query = query, directory = Your_Path,
                      save = TRUE, save.filename = "dataRNA.RData")
## get raw count matrix                         
dataPrep <- TCGAanalyze_Preprocessing(object = dataRNA,
                                      cor.cut = 0.6,
                                      datatype = "HTSeq - Counts")

```

Use `diff_RNA` to do difference analysis. We provide the data of human gene length and GC content in `gene_cov`.

```{r, eval=FALSE, message=FALSE, warning=FALSE}
group <- sample(c("grp1", "grp2"), ncol(dataPrep), replace = TRUE)
library(cqn) # To avoid reporting errors: there is no function "rq"
## get gene length and GC content
library(org.Hs.eg.db)
genes_bitr <- bitr(rownames(gene_cov), fromType = "ENTREZID", toType = "ENSEMBL", 
         OrgDb = org.Hs.eg.db, drop = TRUE)
genes_bitr <- genes_bitr[!duplicated(genes_bitr[,2]), ]
gene_cov2 <- gene_cov[genes_bitr$ENTREZID, ]
rownames(gene_cov2) <- genes_bitr$ENSEMBL
genes <- intersect(rownames(dataPrep), rownames(gene_cov2))
dataPrep <- dataPrep[genes, ]
geneLength <- gene_cov2[genes, "length"]
gccontent <- gene_cov2[genes, "GC"]
names(geneLength) <- names(gccontent) <- genes
##  Difference analysis
DEGAll <- diff_RNA(counts = dataPrep, group = group, 
                   geneLength = geneLength, gccontent = gccontent)
```

Use `clusterProfiler` to do enrichment analytics:

```{r, eval=FALSE, message=FALSE, warning=FALSE}
diffGenes <- DEGAll$logFC
names(diffGenes) <- rownames(DEGAll)
diffGenes <- sort(diffGenes, decreasing = TRUE)
library(clusterProfiler)
library(enrichplot)
library(org.Hs.eg.db)
gsego <- gseGO(gene = diffGenes, OrgDb = org.Hs.eg.db, keyType = "ENSEMBL")
dotplot(gsego)
```



## DNA Methylation data integration 

use `TCGAbiolinks` to download TCGA data

```{r, eval=FALSE, message=FALSE, warning=FALSE}
library(TCGAbiolinks)
query <- GDCquery(project = "TCGA-ACC",
                  data.category = "DNA Methylation",
                  data.type = "Methylation Beta Value",
                  platform = "Illumina Human Methylation 450")
GDCdownload(query, method = "api", files.per.chunk = 5, directory = Your_Path)
```

The function Merge_methy_tcga could Merge methylation data downloaded from TCGA official website or TCGAbiolinks. This makes it easier to extract differentially methylated genes in the downstream analysis. For example:

```{r, eval=FALSE, message=FALSE, warning=FALSE}
merge_result <- Merge_methy_tcga(Your_Path_to_DNA_Methylation_data)
```

Then use `ChAMP` package to do difference analysis.

```{r, eval=FALSE, message=FALSE, warning=FALSE}
library(ChAMP)
diff_gene <- methyDiff(cpgData = merge_result, sampleGroup = sample(c("C","T"), 
    ncol(merge_result[[1]]), replace = TRUE))
```

Use `clusterProfiler` to do enrichment analytics:

```{r, eval=FALSE, message=FALSE, warning=FALSE}
diff_gene$p.adj <- p.adjust(diff_gene$pvalue)
genes <- diff_gene[diff_gene$p.adj < 0.05, "gene"]
library(clusterProfiler)
library(enrichplot)
library(org.Hs.eg.db)
ego <- enrichGO(gene = genes, OrgDb = org.Hs.eg.db, keyType = "SYMBOL")
dotplot(ego)
```


## Copy number variation data integration and differential gene extraction

use TCGAbiolinks to download TCGA data(Gene Level Copy Number Scores)

```{r, eval=FALSE, message=FALSE, warning=FALSE}
library(TCGAbiolinks)
query <- GDCquery(project = "TCGA-LGG",
                  data.category = "Copy Number Variation",
                  data.type = "Gene Level Copy Number Scores")

GDCdownload(query, method = "api", files.per.chunk = 5, directory = Your_Path)

data <- GDCprepare(query = query, 
                   directory =  "Your_Path") 
```

Do difference analysis of gene level copy number variation data using `diff_CNV`

```{r, eval=FALSE, message=FALSE, warning=FALSE}
class(data) <- "data.frame"
cnvData <- data[, -c(1,2,3)]
rownames(cnvData) <- data[, 1]
sampleGroup  = sample(c("A","B"), ncol(cnvData), replace = TRUE)
diffCnv <- diff_CNV(cnvData, sampleGroup)
```

Use `clusterProfiler` to do enrichment analytics:

```{r, eval=FALSE, message=FALSE, warning=FALSE}
pvalues <- diffCnv$pvalue * sign(diffCnv$odds)
genes <- rownames(diffCnv)[diffCnv$pvalue < 0.05]
library(clusterProfiler)
library(enrichplot)
library(org.Hs.eg.db)
ego <- enrichGO(gene = genes, OrgDb = org.Hs.eg.db, keyType = "ENSEMBL")
dotplot(ego)
```

## Difference analysis of single nucleotide Variation data 
Use TCGAbiolinks to download TCGA data

```{r, eval=FALSE, message=FALSE, warning=FALSE}
library(TCGAbiolinks)
query <- GDCquery(project = "TCGA-ACC",
                  data.category = "Simple Nucleotide Variation",
                  data.type = "Masked Somatic Mutation",
                  workflow.type = "MuSE Variant Aggregation and Masking")

GDCdownload(query, method = "api", files.per.chunk = 5, directory = Your_Path)

data_snp <- GDCprepare(query = query, 
                   directory =  "Your_Path") 
```

Use `diff_SNP_tcga` to do difference analysis

```{r, eval=FALSE, message=FALSE, warning=FALSE}
samples <- unique(data_snp$Tumor_Sample_Barcode)
sampleType <- sample(c("A","B"), length(samples), replace = TRUE)
names(sampleType) <- samples
pvalue <- diff_SNP_tcga(snpData = data_snp, sampleType = sampleType)
```

Use `clusterProfiler` to do enrichment analysis

```{r, eval=FALSE, message=FALSE, warning=FALSE}
pvalue2 <- sort(pvalue, decreasing = TRUE)
library(clusterProfiler)
library(enrichplot)
library(org.Hs.eg.db)
gsego <- gseGO(pvalue2, OrgDb = org.Hs.eg.db, keyType = "SYMBOL")
dotplot(gsego)
```


## GEO chip data processing
The function `gene_ave` could average the expression data of different ids for the same gene in the GEO chip data. For example:

```{r, message=FALSE, warning=FALSE}
aa <- c("MARCH1","MARC1","MARCH1","MARCH1","MARCH1")
bb <- c(2.969058399,4.722410064,8.165514853,8.24243893,8.60815086)
cc <- c(3.969058399,5.722410064,7.165514853,6.24243893,7.60815086)
file_gene_ave <- data.frame(aa=aa,bb=bb,cc=cc)
colnames(file_gene_ave) <- c("Gene", "GSM1629982", "GSM1629983")
result <- gene_ave(file_gene_ave, 1)
```

Multiple genes symbols may correspond to a same chip id. The result of function `rep1` is to assign the expression of this id to each gene, and function `rep2` deletes the expression. For example:

```{r}
aa <- c("MARCH1 /// MMA","MARC1","MARCH2 /// MARCH3",
        "MARCH3 /// MARCH4","MARCH1")
bb <- c("2.969058399","4.722410064","8.165514853","8.24243893","8.60815086")
cc <- c("3.969058399","5.722410064","7.165514853","6.24243893","7.60815086")
input_file <- data.frame(aa=aa,bb=bb,cc=cc)
rep1_result <- rep1(input_file," /// ")
rep2_result <- rep2(input_file," /// ")
```

## Other downstream analyses

1. The function `id_conversion_vector` could convert gene id from one of `symbol`, `RefSeq_ID`, `Ensembl_ID`, `NCBI_Gene_ID`, `UCSC_ID`, and `UniProt_ID` , etc. to another. Use `id_ava()` to get all the convertible ids. For example:

```{r}
id_conversion_vector("symbol", "ensembl_gene_id", c("A2ML1", "A2ML1-AS1", "A4GALT", "A12M1", "AAAS")) 
```

When the user converts the Ensembl ID to other ids, the version number needs to be removed. For example, "ENSG00000186092.4" doesn't work, you need to change it to "ENSG00000186092".

Especially, the function `id_conversion` could convert  ENSEMBL gene id to gene Symbol in TCGA. For example:

```{r, eval=FALSE, message=FALSE, warning=FALSE}
profile <- GeoTcgaData::profile
result <- id_conversion(profile)
```

The parameter `profile` is a data.frame or matrix of gene expression data in TCGA.

**Note:** In previous versions(< 1.0.0) the `id_conversion` and `id_conversion_vector` used HGNC data to convert human gene id.  In future versions, we will use `clusterProfiler::bitr` for ID conversion. 

```{r}
library(clusterProfiler)
library(org.Hs.eg.db)
bitr(c("A2ML1", "A2ML1-AS1", "A4GALT", "A12M1", "AAAS"), fromType = "SYMBOL", 
    toType = "ENSEMBL", OrgDb = org.Hs.eg.db, drop = FALSE)
```


2. The function `countToFpkm_matrix` and `countToTpm_matrix` could convert count data to FPKM or TPM data.

```{r}
lung_squ_count2 <- matrix(c(1,2,3,4,5,6,7,8,9),ncol=3)
rownames(lung_squ_count2) <- c("DISC1","TCOF1","SPPL3")
colnames(lung_squ_count2) <- c("sample1","sample2","sample3")
jieguo <- countToFpkm_matrix(lung_squ_count2)
```

```{r, message=FALSE, warning=FALSE}
lung_squ_count2 <- matrix(c(0.11,0.22,0.43,0.14,0.875,0.66,0.77,0.18,0.29),ncol=3)
rownames(lung_squ_count2) <- c("DISC1","TCOF1","SPPL3")
colnames(lung_squ_count2) <- c("sample1","sample2","sample3")
jieguo <- countToTpm_matrix(lung_squ_count2)
```
3. The function `tcga_cli_deal` could combine clinical information obtained from TCGA and extract survival data. For example:

```{r, message=FALSE, warning=FALSE}
tcga_cli <- tcga_cli_deal(system.file(file.path("extdata","tcga_cli"),package="GeoTcgaData"))
```

**Note:** Now the combined clinical data can be downloaded directly from [TCGAbiolinks](http://www.bioconductor.org/packages/release/bioc/vignettes/TCGAbiolinks/inst/doc/clinical.html).
```{r, eval=FALSE, message=FALSE, warning=FALSE}
library(TCGAbiolinks)
## get BCR Biotab data
query <- GDCquery(project = "TCGA-ACC", 
                  data.category = "Clinical",
                  data.type = "Clinical Supplement", 
                  data.format = "BCR Biotab")
GDCdownload(query)
clinical.BCRtab.all <- GDCprepare(query)
names(clinical.BCRtab.all)

## get indexed data
clinical <- GDCquery_clinic(project = "TCGA-ACC", type = "clinical")
```

