---
title: "GeoTcgaData"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GeoTcgaData}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```
--------

## Authors
Erqiang Hu

Department of Bioinformatics, School of Basic Medical Sciences, 
Southern Medical University.


## Introduction
GEO and TCGA provide us with a wealth of data, such as RNA-seq, DNA Methylation,
single nucleotide Variation and Copy number variation data. 
It's easy to download data from TCGA using the  gdc tool or `TCGAbiolinks`, 
and some software provides organized TCGA data, such as 
[UCSC Xena](http://xena.ucsc.edu/) , UCSCXenaTools, and 
[sangerbox](http://vip.sangerbox.com/), but processing these data into a format
suitable for bioinformatics  analysis requires more work. This R package was 
developed to handle these data.

```{r setup}
library(GeoTcgaData)
```

## Example

This is a basic example which shows you how to solve a common problem:

### RNA-seq data differential expression analysis

It is convenient to use TCGAbiolinks  or 
[`GDCRNATools`](https://bioconductor.org/packages/GDCRNATools/) to download 
and analysis Gene expression data.  `TCGAbiolinks` use `edgeR` package to do 
differential expression analysis, while `GDCRNATools` can implement three most 
commonly used methods: limma, edgeR , and DESeq2 to identify differentially 
expressed  genes (DEGs).

Alicia Oshlack  et al. claimed that unlike the chip data, 
the RNA-seq data had one [bias](https://pubmed.ncbi.nlm.nih.gov/20132535/)[1]: 
the larger the transcript length / mean read count , the more likely it was to 
be  identified as a differential gene, 
while there was no such trend in the 
[chip data](https://pubmed.ncbi.nlm.nih.gov/19371405/)[2].


 However, when we use their chip data for difference analysis
 (using the limma package), we find that chip data has the same trend as 
 RNA-seq data. And we also found this trend in the difference analysis results 
 given by the data 
 [authors](https://genome.cshlp.org/content/18/9/1509.long)[3].

 

 It is worse noting that only technical replicate data, which has small gene 
 dispersions, shows this [bias](https://pubmed.ncbi.nlm.nih.gov/28545404/)[4]. 
 This is because in technical replicate RNA-seq data a long gene has more 
 reads mapping to it compared to a short gene of similar expression, 
 and most of the statistical methods used to detect differential expression
 have stronger detection ability for genes with more reads. However, we have 
 not deduced why there is such a bias in the current difference 
 analysis algorithms. 

Some software, such as [CQN](http://www.bioconductor.org/packages/cqn/) ,
present a 
[normalization algorithm](https://pubmed.ncbi.nlm.nih.gov/22285995/) [5]
to correct systematic biases(gene length bias and 
[GC-content bias](https://pubmed.ncbi.nlm.nih.gov/22177264/)[6]. 
But they did not provide sufficient evidence to prove that the correction is 
effective. We use the 
[Marioni dataset](https://pubmed.ncbi.nlm.nih.gov/19371405/)[2] to verify the
correction effect of CQN and find that there is still a deviation 
after correction:



[GOseq](http://bioconductor.org/packages/goseq/) [1]based on 
Wallenius' noncentral hypergeometric distribution can effectively correct the 
gene length deviation in enrichment analysis. However, the current RNA-seq data 
often have no gene length bias, but only the expression amount(read count) 
bias, GOseq may overcorrect these data, correcting originally unbiased data 
into reverse bias.

GOseq also fails to correct for expression bias, therefore, read count bias 
correction is still a challenge for us.



use `TCGAbiolinks` data:  

```{r, eval=FALSE, message=FALSE, warning=FALSE}
# use `TCGAbiolinks` to download TCGA data
library(TCGAbiolinks)
                
query <- GDCquery(project = "TCGA-ACC",
                data.category = "Transcriptome Profiling",
                data.type = "Gene Expression Quantification", 
                workflow.type = "STAR - Counts")
                
GDCdownload(query, method = "api", files.per.chunk = 3, 
    directory = Your_Path)

dataRNA <- GDCprepare(query = query, directory = Your_Path,
                    save = TRUE, save.filename = "dataRNA.RData")
## get raw count matrix                         
dataPrep <- TCGAanalyze_Preprocessing(object = dataRNA,
    cor.cut = 0.6, datatype = "STAR - Counts")


# Use `diff_RNA` to do difference analysis. We provide the data 
# of human gene length and GC content in `gene_cov`.
group <- sample(c("grp1", "grp2"), ncol(dataPrep), replace = TRUE)
library(cqn) # To avoid reporting errors: there is no function "rq"
## get gene length and GC content
library(org.Hs.eg.db)
genes_bitr <- bitr(rownames(gene_cov), fromType = "ENTREZID", 
    toType = "ENSEMBL", 
    OrgDb = org.Hs.eg.db, drop = TRUE)
genes_bitr <- genes_bitr[!duplicated(genes_bitr[,2]), ]
gene_cov2 <- gene_cov[genes_bitr$ENTREZID, ]
rownames(gene_cov2) <- genes_bitr$ENSEMBL
genes <- intersect(rownames(dataPrep), rownames(gene_cov2))
dataPrep <- dataPrep[genes, ]
geneLength <- gene_cov2[genes, "length"]
gccontent <- gene_cov2[genes, "GC"]
names(geneLength) <- names(gccontent) <- genes
##  Difference analysis
DEGAll <- diff_RNA(counts = dataPrep, group = group, 
    geneLength = geneLength, gccontent = gccontent)

# Use `clusterProfiler` to do enrichment analytics:
diffGenes <- DEGAll$logFC
names(diffGenes) <- rownames(DEGAll)
diffGenes <- sort(diffGenes, decreasing = TRUE)
library(clusterProfiler)
library(enrichplot)
library(org.Hs.eg.db)
gsego <- gseGO(gene = diffGenes, OrgDb = org.Hs.eg.db, keyType = "ENSEMBL")
dotplot(gsego)
```


### DNA Methylation data integration 

use `TCGAbiolinks` data. 

The codes may need to be modified if `TCGAbiolinks` updates. 
So please read its documents.

```{r, eval=FALSE, message=FALSE, warning=FALSE}
# download data
library(TCGAbiolinks)
query <- GDCquery(project = "TCGA-ACC",
    data.category = "DNA Methylation",
    data.type = "Methylation Beta Value",
    platform = "Illumina Human Methylation 450")
GDCdownload(query, method = "api", files.per.chunk = 5, directory = Your_Path)

# The function Merge_methy_tcga could Merge methylation data
# downloaded from TCGA official website or TCGAbiolinks. 
# This makes it easier to extract differentially methylated
# genes in the downstream analysis. For example:
merge_result <- Merge_methy_tcga(Your_Path_to_DNA_Methylation_data)

# Then use `methyDiff()` to do difference analysis.
library(ChAMP) # To avoid reporting errors
diff_gene <- methyDiff(cpgData = merge_result, sampleGroup = sample(c("C","T"), 
    ncol(merge_result[[1]]), replace = TRUE))

# use `methyDiff_ucsc` to do difference analysis   
methy_file <- "TCGA.THCA.sampleMap_HumanMethylation450.gz"
methy <- fread(methy_file, sep = "\t", header = T)
library(ChAMP)
myImport <- champ.import(directory=system.file("extdata",package="ChAMPdata"))
myfilter <- champ.filter(beta=myImport$beta,pd=myImport$pd,
    detP=myImport$detP,beadcount=myImport$beadcount)
cpg_gene <- hm450.manifest.hg19[, c("probeID", "gene_HGNC")]
## or use IlluminaHumanMethylation450kanno.ilmn12.hg19 to get annotation data
library(IlluminaHumanMethylation450kanno.ilmn12.hg19)
# ann <- getAnnotation(IlluminaHumanMethylation450kanno.ilmn12.hg19)
# class(ann) <- "data.frame"
# cpg_gene <- ann[,c("Name", "UCSC_RefGene_Name", "UCSC_RefGene_Group")]

methy_df <- methyDiff_ucsc(methy, cpg_gene)
# Use `clusterProfiler` to do enrichment analytics:
diff_gene$p.adj <- p.adjust(diff_gene$pvalue)
genes <- diff_gene[diff_gene$p.adj < 0.05, "gene"]
library(clusterProfiler)
library(enrichplot)
library(org.Hs.eg.db)
ego <- enrichGO(gene = genes, OrgDb = org.Hs.eg.db, keyType = "SYMBOL")
dotplot(ego)
```
**Note:** `ChAMP`has a large number of dependent packages.
If you cannot install it  successfully, you can download each dependent package 
separately(Source or Binary) and install it  locally.

We provide two models to get methylation difference genes:  

if model = "cpg", step1: calculate difference cpgs; 
step2: calculate difference genes; 

if model = "gene", step1: calculate the methylation level of genes;
step2: calculate difference genes.

We find that only model = "gene" has no deviation of CpG number. 


### Copy number variation data integration and differential gene extraction

use TCGAbiolinks to download TCGA data(Gene Level Copy Number Scores)

```{r, eval=FALSE, message=FALSE, warning=FALSE}
# download data
library(TCGAbiolinks)
query <- GDCquery(project = "TCGA-LGG",
        data.category = "Copy Number Variation",
        data.type = "Gene Level Copy Number Scores")

GDCdownload(query, method = "api", files.per.chunk = 5, directory = Your_Path)

data <- GDCprepare(query = query, 
    directory =  "Your_Path") 
# Do difference analysis of gene level copy number 
# variation data using `diff_CNV`
class(data) <- "data.frame"
cnvData <- data[, -c(1,2,3)]
rownames(cnvData) <- data[, 1]
sampleGroup  <- sample(c("A","B"), ncol(cnvData), replace = TRUE)
diffCnv <- diff_CNV(cnvData, sampleGroup)

# Use `clusterProfiler` to do enrichment analytics:
pvalues <- diffCnv$pvalue * sign(diffCnv$odds)
genes <- rownames(diffCnv)[diffCnv$pvalue < 0.05]
library(clusterProfiler)
library(enrichplot)
library(org.Hs.eg.db)
ego <- enrichGO(gene = genes, OrgDb = org.Hs.eg.db, keyType = "ENSEMBL")
dotplot(ego)
```



### Difference analysis of single nucleotide Variation data 

```{r, eval=FALSE, message=FALSE, warning=FALSE}
# download data
library(TCGAbiolinks)
query <- GDCquery(project = "TCGA-ACC",
    data.category = "Simple Nucleotide Variation",
    data.type = "Masked Somatic Mutation",
    workflow.type = "MuSE Variant Aggregation and Masking")

GDCdownload(query, method = "api", files.per.chunk = 5, directory = Your_Path)

data_snp <- GDCprepare(query = query, 
    directory =  "Your_Path") 
# Use `diff_SNP_tcga` to do difference analysis
samples <- unique(data_snp$Tumor_Sample_Barcode)
sampleType <- sample(c("A","B"), length(samples), replace = TRUE)
names(sampleType) <- samples
pvalue <- diff_SNP_tcga(snpData = data_snp, sampleType = sampleType)

# Use `clusterProfiler` to do enrichment analysis
pvalue2 <- sort(pvalue, decreasing = TRUE)
library(clusterProfiler)
library(enrichplot)
library(org.Hs.eg.db)
gsego <- gseGO(pvalue2, OrgDb = org.Hs.eg.db, keyType = "SYMBOL")
dotplot(gsego)
```


### GEO chip data processing
The function `gene_ave` could average the expression data of different 
ids for the same gene in the GEO chip data. For example:

```{r, message=FALSE, warning=FALSE}
aa <- c("MARCH1","MARC1","MARCH1","MARCH1","MARCH1")
bb <- c(2.969058399,4.722410064,8.165514853,8.24243893,8.60815086)
cc <- c(3.969058399,5.722410064,7.165514853,6.24243893,7.60815086)
file_gene_ave <- data.frame(aa=aa,bb=bb,cc=cc)
colnames(file_gene_ave) <- c("Gene", "GSM1629982", "GSM1629983")
result <- gene_ave(file_gene_ave, 1)
```

Multiple genes symbols may correspond to a same chip id. The result of 
function `rep1` is to assign the expression of this id to each gene, 
and function `rep2` deletes the expression. For example:

```{r}
aa <- c("MARCH1 /// MMA","MARC1","MARCH2 /// MARCH3",
        "MARCH3 /// MARCH4","MARCH1")
bb <- c("2.969058399","4.722410064","8.165514853","8.24243893","8.60815086")
cc <- c("3.969058399","5.722410064","7.165514853","6.24243893","7.60815086")
input_file <- data.frame(aa=aa,bb=bb,cc=cc)
rep1_result <- rep1(input_file," /// ")
rep2_result <- rep2(input_file," /// ")
```

### Other downstream analyses

1. The function `id_conversion_vector` could convert gene id from one 
of `symbol`, `RefSeq_ID`, `Ensembl_ID`, `NCBI_Gene_ID`, `UCSC_ID`, 
and `UniProt_ID` , etc. to another. Use `id_ava()` to get all the 
convertible ids. For example:

```{r}
id_conversion_vector("symbol", "ensembl_gene_id", c("A2ML1", "A2ML1-AS1", 
    "A4GALT", "A12M1", "AAAS")) 
```

When the user converts the Ensembl ID to other ids, the version number needs 
to be removed. For example, "ENSG00000186092.4" doesn't work, you need to 
change it to "ENSG00000186092".

Especially, the function `id_conversion` could convert  ENSEMBL gene id to 
gene Symbol in TCGA. For example:

```{r, message=FALSE, warning=FALSE}
data(profile)
result <- id_conversion(profile)
```

The parameter `profile` is a data.frame or matrix of gene expression 
data in TCGA.

**Note:** In previous versions(< 1.0.0) the `id_conversion` and 
`id_conversion_vector` used HGNC data to convert human gene id.  
In future versions, we will use `clusterProfiler::bitr` for ID conversion. 

```{r}
library(clusterProfiler)
library(org.Hs.eg.db)
bitr(c("A2ML1", "A2ML1-AS1", "A4GALT", "A12M1", "AAAS"), fromType = "SYMBOL", 
    toType = "ENSEMBL", OrgDb = org.Hs.eg.db, drop = FALSE)
```


2. The function `countToFpkm_matrix` and `countToTpm_matrix` could convert 
count data to FPKM or TPM data.

```{r}
data(gene_cov)
lung_squ_count2 <- matrix(c(1,2,3,4,5,6,7,8,9),ncol=3)
rownames(lung_squ_count2) <- c("DISC1","TCOF1","SPPL3")
colnames(lung_squ_count2) <- c("sample1","sample2","sample3")
result <- countToFpkm_matrix(lung_squ_count2, keyType = "SYMBOL", 
    gene_cov = gene_cov)
```

```{r, message=FALSE, warning=FALSE}
data(gene_cov)
lung_squ_count2 <- matrix(c(0.11,0.22,0.43,0.14,0.875,
    0.66,0.77,0.18,0.29),ncol=3)
rownames(lung_squ_count2) <- c("DISC1","TCOF1","SPPL3")
colnames(lung_squ_count2) <- c("sample1","sample2","sample3")
result <- countToTpm_matrix(lung_squ_count2, keyType = "SYMBOL", 
    gene_cov = gene_cov)
```
3. The function `tcga_cli_deal` could combine clinical information obtained 
from TCGA and extract survival data. For example:

```{r, message=FALSE, warning=FALSE}
tcga_cli <- tcga_cli_deal(system.file(file.path("extdata","tcga_cli"),
    package="GeoTcgaData"))
```

**Note:** Now the combined clinical data can be downloaded 
directly from TCGAbiolinks.
```{r, eval=FALSE, message=FALSE, warning=FALSE}
library(TCGAbiolinks)
## get BCR Biotab data
query <- GDCquery(project = "TCGA-ACC", 
    data.category = "Clinical",
    data.type = "Clinical Supplement", 
    data.format = "BCR Biotab")
GDCdownload(query)
clinical.BCRtab.all <- GDCprepare(query)
names(clinical.BCRtab.all)

## get indexed data
clinical <- GDCquery_clinic(project = "TCGA-ACC", type = "clinical")
```

```{r}
sessionInfo()
```

## References
1. Young MD, Wakefield MJ, Smyth GK, Oshlack A (2010) Gene ontology analysis 
for RNA-seq: accounting for selection bias. Genome Biol 11: R14.
2. Oshlack A, Wakefield MJ (2009) Transcript length bias in RNA-seq data 
confounds systems biology. Biol Direct 4: 14.
3. Marioni JC, Mason CE, Mane SM, Stephens M, Gilad Y (2008) RNA-seq: an 
assessment of technical reproducibility and comparison with gene expression 
arrays. Genome Res 18: 1509-1517.
4. Yoon S, Nam D (2017) Gene dispersion is the key determinant of the read 
count bias in differential expression analysis of RNA-seq data. 
BMC Genomics 18: 408.
5. Hansen KD, Irizarry RA, Wu Z (2012) Removing technical variability in 
RNA-seq data using conditional quantile normalization. 
Biostatistics 13: 204-216.
6. Risso D, Schwartz K, Sherlock G, Dudoit S (2011) GC-content normalization
for RNA-Seq data. BMC Bioinformatics 12: 480.

